CrypTool Script \\ Mathematics and Cryptography

### 2002-07-12 script/en/chapters/primes.tex
misc
be remarks

%\paragraph{What are prime numbers?}
\subsection*{What are prime numbers?}
\addcontentsline{toc}{subsection}{~~~~~~What are prime numbers?}
\index{Prime number} \index{Numbers!prime numbers}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

The {\em natural} \index{Numbers} numbers $\mathbb{N}=\{1, 2, 3, 4,\cdots \}$ thus comprise 
\begin{itemize}
   \item the number $1$ (the unit value)
   \item the primes and
   \item the composite numbers.
\end{itemize}

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public 
key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signatures!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance of computers) and leads
to new calculation methods on many computers \\ (see also:
\href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millenniums.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also explain
what is known about the distribution (density, number of prime numbers in
particular intervals) of prime numbers and how prime number tests work.

@@

\begin{definition}\label{def-pz-prime}
A whole number $p \in {\bf N}$ is called prime \index{Numbers!prime numbers} if
$p > 1$ and $p$ only possesses the trivial factors $\pm 1$ and $\pm p$.
\end{definition}

@@

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this, the percentage of primes
constantly decreases. Prime numbers can be factorised in a unique {\em trivial}
way: $$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors are called \index{Numbers!composite}
{\em composite}. These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

\index{Numbers!Mersenne}\index{Mersenne!number}
\index{Mersenne!theorem} Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). A prime Mersenne number \index{Mersenne!prime number} is called Mersenne prime number.  
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

@@

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne!number!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne!number!M-39}

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (small) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!small Fermat}
\begin{theorem}[``small'' Fermat]\label{thm-pz-fermat1}
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime numbers}\index{Numbers!pseudoprime
number} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Numbers!Carmichael numbers} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

Detailed descriptions of tests for finding our whether a number is prime can be
found on Web sites such as:

@@

  \item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circel division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Numbers!Fermat numbers}\index{Fermat!Fermat
number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned above, Fermat wrote to Mersenne regarding this assumption.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65,537 &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 ,294, 967, 297 = 641 \cdot
6,700,417 &  \mapsto {\rm ~NOT~prime} !
\end{array} 
$$
  \item Carmichael numbers: see above.
  \item Pseudoprime numbers: see above.
  \item Strong pseudoprime numbers: see above.
  \item Idea based on \hyperlink{euclid}{Euclid's proof} (infinite many prime
numbers) $p_1 \cdot p_2 \cdots p_n +1$:

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about all these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime number!relative} {\em relatively prime }.
  \item $f(n) = n^2 + n + 41$:\\
  This sequence starts off very {\em promisingly},   but is far from being a
proof.

@@

The first $40$ values are prime numbers (which have the obvious regularity that
their difference starts with $2$ and increases by $2$ each time), but the $41$th
and $42$th values are not prime numbers.     It is easy to see that $f(41)$ hl
cannot be a prime number:
    $f(41) = 41^2 + 41 + 41 = 41 (41 + 1 + 1) = 41 \cdot 43$.
  \item $f(n) = n^2 - 79 \cdot n + 1,601$: \\
    This function delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
difference prime values (the same ones as the function from point 10).
$$
\begin{array}{|ll||ll|}
\hline
f(0) = 1,601    & \mapsto {\rm ~prime} &  f(28) = 173    & \mapsto {\rm ~prime}
\\
f(1) = 1,523    & \mapsto {\rm ~prime} &  f(29) = 151    & \mapsto {\rm ~prime}
\\
f(2) = 1,447    & \mapsto {\rm ~prime} &  f(30) = 131 & \mapsto {\rm ~prime} \\
f(3) = 1,373    & \mapsto {\rm ~prime} &  f(31) = 113 & \mapsto {\rm ~prime} \\
f(4) = 1,301    & \mapsto {\rm ~prime} &  f(32) = 97 & \mapsto {\rm ~prime} \\
f(5) = 1,231    & \mapsto {\rm ~prime} &  f(33) = 83 & \mapsto {\rm ~prime} \\
f(6) = 1,163    & \mapsto {\rm ~prime} &  f(34) = 71 & \mapsto {\rm ~prime} \\
f(7) = 1,097    & \mapsto {\rm ~prime} &  f(35) = 61 & \mapsto {\rm ~prime} \\
f(8) = 1,033    & \mapsto {\rm ~prime} &  f(36) = 53 & \mapsto {\rm ~prime} \\
f(9) = 971  & \mapsto {\rm ~prime} &  f(37) = 47 & \mapsto {\rm ~prime} \\
f(10) = 911 & \mapsto {\rm ~prime} &  f(38) = 43 & \mapsto {\rm ~prime} \\
f(11) = 853 & \mapsto {\rm ~prime} &  f(39) = 41 & \mapsto {\rm ~prime} \\
f(12) = 797 & \mapsto {\rm ~prime} &             &                     \\
f(13) = 743 & \mapsto {\rm ~prime} &  f(40) = 41 & \mapsto {\rm ~prime} \\
f(14) = 691 & \mapsto {\rm ~prime} &  f(41) = 43 & \mapsto {\rm ~prime} \\
f(15) = 641 & \mapsto {\rm ~prime} &  f(42) = 47 & \mapsto {\rm ~prime} \\
f(16) = 593 & \mapsto {\rm ~prime} &  f(43) = 53 & \mapsto {\rm ~prime} \\
f(17) = 547 & \mapsto {\rm ~prime} &  \cdots  &  \\
f(18) = 503 & \mapsto {\rm ~prime} &  f(77) = 1,447  & \mapsto {\rm ~prime} \\
f(19) = 461 & \mapsto {\rm ~prime} &  f(78) = 1,523  & \mapsto {\rm ~prime} \\
f(20) = 421 & \mapsto {\rm ~prime} &  f(79) = 1,601  & \mapsto {\rm ~prime} \\
f(21) = 383 & \mapsto {\rm ~prime} &  f(80) = 11 \cdot 151 & \mapsto {\rm
~NOT~prime!} \\
f(22) = 347 & \mapsto {\rm ~prime} &  f(81) = 41 \cdot 43 & \mapsto {\rm
~NOT~prime!} \\
f(21) = 383 & \mapsto {\rm ~prime} &  f(82) = 1,847  & \mapsto {\rm ~prime} \\
f(22) = 347 & \mapsto {\rm ~prime} &  f(83) = 1,933  & \mapsto {\rm ~prime} \\
f(23) = 313 & \mapsto {\rm ~prime} &  f(84) = 43 \cdot 47 &  \mapsto {\rm
~NOT~prime!} \\
f(24) = 281 & \mapsto {\rm ~prime} & & \\
f(25) = 251 & \mapsto {\rm ~prime} & & \\
f(26) = 223 & \mapsto {\rm ~prime} & & \\
f(27) = 197 & \mapsto {\rm ~prime} & & \\
\hline
\end{array} 
$$
  \item Polynomial functions $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 +
a_0$  ($a_i$ in ${\mathbb Z}$, $n \geq 1$):

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
(1690-1764) conjectured:
    Every even natural number greater than 2 can be represented as the sum of
two prime numbers.
    Computers have verified\footnote{It is generally accepted today, that
the Goldbach Conjecture is true, i.e. valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richtstein from the computer sciences institute at the University
of Giessen, studied even numbers up to 400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact that despite all efforts, Goldbachs' conjecture has to date not been proven leads
one to believe that: since the pioneer work of the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is well-known,
not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbachs' Conjecture was correct, despite that however,
proof will never be found. Conversely, that will presumably also remain unproven.}
the Goldbach conjecture for all even numbers up to
$4*10^{14}$ but no general proof has yet been found\footnote{The English
publisher {\em Faber} and the American publisher {\em Bloomsbury} issued in 2000
the 1992 published book ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than 250 old puzzle. To
stimulate the sell rates the English and American publisher compete a prize of 1 million USD, 
is someone proves the conjecture -- to be published till 2004 by an well-known mathematical journal.\\
Susprisingly only British and American citizens are allowed to participate.\\
The theorem closest to Goldbach's conjecture yet was proven 1966 by Chen Jing-Run in a hard to understand fashion:
Each even integer bigger than $2$ is the sum of one prime and of the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.}.
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
(1826-1866) put forward a formula for the distribution of primes that would
further improve the estimate. However, this has neither been proved nor
contradicted so far.
\end{itemize}

@@

  \item Does a polynomial time algorithm exist for calculating the prime factors of a number (see
    \cite[p. 167]{Klee1997})? 
    This question can be divided into the two following questions:
\begin{itemize} 
 \item Does a polynomial time algorithm exist that decides whether a number is
prime?
  \item Does a polynomial time algorithm exist that for a composite number $n$
calculates a non-trivial (i.e. other than $1$ and $n$) factor of $n$?
\end{itemize}
\end{itemize}

@@

\begin{tabular}{|l|l|l|l|}\hline
Index   &   Precise value  & Rounded value &
Comment \\
\hline \hline
1       &   2             &     2              & \\
2       &   3             &     3  &  \\
3       &   5             &     5  & \\
4       &   7             &     7 & \\
5       &   11            &     11 & \\
6       &   13            &     13 & \\
7       &   17            &     17 & \\
8       &   19            &     19 & \\
9       &   23            &     23 & \\
10      &   29            &     29 & \\
100     &   541           &     541 & \\
1,000    &   7,917          &     7,917 & \\
664,559  &  9,999,991     &     9.99999E+06 &   All prime numbers up to 1E+07 were known\\
        &                 &                 &  at the beginning of the 20th century. \\
1E+06  &    15,485,863   &      1.54859E+07 & \\
6E+06  &    104,395,301    &    1.04395E+08  & This prime was discovered in 1959.\\
1E+07  &    179,424,673     &    1.79425E+08 & \\
1E+09  &    22,801,763,489  &    2.28018E+10 & \\
1E+12  &    29,996,224,275,833 & 2.99962E+13 & \\ \hline
\end{tabular}
\vskip +2 pt Comment: With gaps, extremely large prime numbers were discovered
at an early stage.  \\
\vskip +10pt URLs:

>>>>

\begin{center}
\fbox{\parbox{15cm}{
    \emph{Albert Einstein\footnotemark:}\\
    Progress requires exchange of knowledge.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  German physicist and Nobel Prize winner, March 14, 1879 $-$ April 14, 1955
}

%\paragraph{What are prime numbers?}
\subsection*{What are prime numbers?}
\addcontentsline{toc}{subsection}{~~~~~~What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

The {\em natural} \index{Number!natural} numbers $\mathbb{N}=\{1, 2, 3, 4,\cdots \}$ thus comprise 
\begin{itemize}
   \item the number $1$ (the unit value)
   \item the primes and
   \item the composite numbers.
\end{itemize}

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signatures!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\ (see also:
\href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millenniums.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also explain
what is known about the distribution (density, number of prime numbers in
particular intervals) of prime numbers and how prime number tests work.

@@

\begin{definition}\label{def-pz-prime}
A whole number $p \in {\bf N}$ is called prime \index{Number!prime} if
$p > 1$ and $p$ only possesses the trivial factors $\pm 1$ and $\pm p$.
\end{definition}

@@

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this, the percentage of primes
constantly decreases. Prime numbers can be factorised in a unique {\em trivial}
way: $$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors are called \index{Number!composite}
{\em composite}. These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

\index{Number!Mersenne}\index{Mersenne!number}
\index{Mersenne!theorem} Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). A prime Mersenne number \index{Mersenne!prime number} is called Mersenne prime number.  
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

@@

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne!prime number!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne!prime number!M-39}

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (little) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!little theorem}
\begin{theorem}[``little'' Fermat]\label{thm-pz-fermat1}
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime numbers}\index{Number!pseudoprime} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Number!Carmichael} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

Detailed descriptions of tests for finding out whether a number is prime can be
found on Web sites such as:

@@

  \item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circel division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned above, Fermat wrote to Mersenne regarding this assumption.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65,537 &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 ,294, 967, 297 = 641 \cdot
6,700,417 &  \mapsto {\rm ~NOT~prime} !
\end{array} 
$$
  \item Carmichael numbers: see above.
  \item Pseudoprime numbers: see above.
  \item Strong pseudoprime numbers: see above.
  \item Idea based on \hyperlink{euclid}{Euclid's proof} (infinite many prime
numbers) $p_1 \cdot p_2 \cdots p_n +1$:

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about all these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime number!relative} {\em relatively prime}.
\pagebreak
  \item $f(n) = n^2 + n + 41$:\\
  This sequence starts off very {\em promisingly},   but is far from being a
proof.

@@

The first $40$ values are prime numbers (which have the obvious regularity that
their difference starts with $2$ and increases by $2$ each time), but the $41$th
and $42$th values are not prime numbers.     It is easy to see that $f(41)$ hl
cannot be a prime number:
    $f(41) = 41^2 + 41 + 41 = 41 (41 + 1 + 1) = 41 \cdot 43$.
\pagebreak
  \item $f(n) = n^2 - 79 \cdot n + 1,601$: \\
    This function delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
difference prime values (the same ones as the function from point 10).
$$
\begin{array}{|ll||ll|}
\hline
f(0) = 1,601    & \mapsto {\rm ~prime} &  f(28) = 173    & \mapsto {\rm ~prime}
\\
f(1) = 1,523    & \mapsto {\rm ~prime} &  f(29) = 151    & \mapsto {\rm ~prime}
\\
f(2) = 1,447    & \mapsto {\rm ~prime} &  f(30) = 131 & \mapsto {\rm ~prime} \\
f(3) = 1,373    & \mapsto {\rm ~prime} &  f(31) = 113 & \mapsto {\rm ~prime} \\
f(4) = 1,301    & \mapsto {\rm ~prime} &  f(32) = 97 & \mapsto {\rm ~prime} \\
f(5) = 1,231    & \mapsto {\rm ~prime} &  f(33) = 83 & \mapsto {\rm ~prime} \\
f(6) = 1,163    & \mapsto {\rm ~prime} &  f(34) = 71 & \mapsto {\rm ~prime} \\
f(7) = 1,097    & \mapsto {\rm ~prime} &  f(35) = 61 & \mapsto {\rm ~prime} \\
f(8) = 1,033    & \mapsto {\rm ~prime} &  f(36) = 53 & \mapsto {\rm ~prime} \\
f(9) = 971  & \mapsto {\rm ~prime} &  f(37) = 47 & \mapsto {\rm ~prime} \\
f(10) = 911 & \mapsto {\rm ~prime} &  f(38) = 43 & \mapsto {\rm ~prime} \\
f(11) = 853 & \mapsto {\rm ~prime} &  f(39) = 41 & \mapsto {\rm ~prime} \\
f(12) = 797 & \mapsto {\rm ~prime} &             &                     \\
f(13) = 743 & \mapsto {\rm ~prime} &  f(40) = 41 & \mapsto {\rm ~prime} \\
f(14) = 691 & \mapsto {\rm ~prime} &  f(41) = 43 & \mapsto {\rm ~prime} \\
f(15) = 641 & \mapsto {\rm ~prime} &  f(42) = 47 & \mapsto {\rm ~prime} \\
f(16) = 593 & \mapsto {\rm ~prime} &  f(43) = 53 & \mapsto {\rm ~prime} \\
f(17) = 547 & \mapsto {\rm ~prime} &  \cdots  &  \\
f(18) = 503 & \mapsto {\rm ~prime} &  f(77) = 1,447  & \mapsto {\rm ~prime} \\
f(19) = 461 & \mapsto {\rm ~prime} &  f(78) = 1,523  & \mapsto {\rm ~prime} \\
f(20) = 421 & \mapsto {\rm ~prime} &  f(79) = 1,601  & \mapsto {\rm ~prime} \\
f(21) = 383 & \mapsto {\rm ~prime} &  f(80) = 11 \cdot 151 & \mapsto {\rm
~NOT~prime!} \\
f(22) = 347 & \mapsto {\rm ~prime} &  f(81) = 41 \cdot 43 & \mapsto {\rm
~NOT~prime!} \\
f(21) = 383 & \mapsto {\rm ~prime} &  f(82) = 1,847  & \mapsto {\rm ~prime} \\
f(22) = 347 & \mapsto {\rm ~prime} &  f(83) = 1,933  & \mapsto {\rm ~prime} \\
f(23) = 313 & \mapsto {\rm ~prime} &  f(84) = 43 \cdot 47 &  \mapsto {\rm
~NOT~prime!} \\
f(24) = 281 & \mapsto {\rm ~prime} & & \\
f(25) = 251 & \mapsto {\rm ~prime} & & \\
f(26) = 223 & \mapsto {\rm ~prime} & & \\
f(27) = 197 & \mapsto {\rm ~prime} & & \\
\hline
\end{array} 
$$
  \item Polynomial functions $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 +
a_0$  ($a_i$ in ${\mathbb Z}$, $n \geq 1$):

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
(1690-1764) conjectured:
    Every even natural number greater than 2 can be represented as the sum of
two prime numbers.
    Computers have verified\footnote{It is generally accepted today, that
the Goldbach Conjecture is true, i.~e.\ valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richtstein from the computer sciences institute at the University
of Giessen, studied even numbers up to 400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact that despite all efforts, Goldbachs' conjecture has to date not been proven leads
one to believe that: since the pioneer work of the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is well-known,
not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbachs' Conjecture was correct, despite that however,
proof will never be found. Conversely, that will presumably also remain unproven.}
the Goldbach conjecture for all even numbers up to
$4*10^{14}$ but no general proof has yet been found\footnote{The English
publisher {\em Faber} and the American publisher {\em Bloomsbury} issued in 2000
the 1992 published book ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than 250
year old puzzle. To
stimulate the sell rates the English and American publishers compete a prize of 1 million USD, 
if someone proves the conjecture -- to be published till 2004 by a well-known mathematical journal.\\
Susprisingly only British and American citizens are allowed to participate.\\
The theorem closest to Goldbach's conjecture yet was proven 1966 by Chen Jing-Run in a hard to understand fashion:
Each even integer bigger than $2$ is the sum of one prime and of the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes transparent, that even today we do not
have fully understood the deeper connections between addition and
multiplication of natural numbers.}.
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
(1826-1866) put forward a formula for the distribution of primes that would
further improve the estimate. However, this has neither been proved nor
contradicted so far.
\end{itemize}

@@

  \item Does a polynomial time algorithm exist for calculating the prime factors of a number (see
    \cite[p. 167]{Klee1997})? 
    This question can be divided into the two following questions:
\begin{itemize} 
 \item Does a polynomial time algorithm exist that decides whether a number is
prime?
\item Does a polynomial time algorithm exist that calculates for a
  composite number $n$ a non-trivial (i.e. other than $1$ and $n$) factor
  of $n$?\footnote{Please see chapters \ref{RSABernstein} and
    \ref{NoteFactorisation}.}
\end{itemize}
\end{itemize}

@@

\begin{tabular}{|l|l|l|l|}\hline
Index   &   Precise value  & Rounded value &
Comment \\
\hline \hline
1       &   2             &     2              & \\
2       &   3             &     3  &  \\
3       &   5             &     5  & \\
4       &   7             &     7 & \\
5       &   11            &     11 & \\
6       &   13            &     13 & \\
7       &   17            &     17 & \\
8       &   19            &     19 & \\
9       &   23            &     23 & \\
10      &   29            &     29 & \\
100     &   541           &     541 & \\
1,000    &   7,917          &     7,917 & \\
664,559  &  9,999,991     &     9.99999E+06 &   All prime numbers up to 1E+07 were known\\
        &                 &                 &  at the beginning of the 20th century. \\
1E+06  &    15,485,863   &      1.54859E+07 & \\
6E+06  &    104,395,301    &    1.04395E+08  & This prime was discovered in 1959.\\
1E+07  &    179,424,673     &    1.79425E+08 & \\
1E+09  &    22,801,763,489  &    2.28018E+10 & \\
1E+12  &    29,996,224,275,833 & 2.99962E+13 & \\ \hline
\end{tabular}
\vskip +2 pt Comment: With gaps, extremely large prime numbers were discovered
at an early stage.  \\
\vskip +10pt Web links:
